---
title: 'tidytuesday #48: Student Loan Debt'
author: '@SchackartK'
output:
  html_document: default
  pdf_document: default
---
# The Data

Data represented here originally come from the US Department of Education. These data were fetched from the [tidytuesday Github repo](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-11-26). Information is provided for several financing agencies for student loans in the United States. The information used is the `Total Payment`s to each lending `Agency` (when available) for the fiscal years 2016-2018. This total is broken down into 4 categories:

* `Consolidation`
* `Rehabilitation`
* `Voluntary Payments`
* `Wage Garnishments`

# Visualization

## Agency Category Breakdown

Here is how the `Total Repayment` for each `Agency` breaks down by `Repayment Type`, averaged over all years that data were present for that `Agency`.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.width=10}
library(tidyverse)
library(viridis)
library(GGally)
library(kableExtra)

# Get the dataset
loans <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-26/loans.csv")

#Clean up the agency names to be uniform
loans$agency_name[grepl("ACT",loans$agency_name)]<-"Account Control"

for(agency_i in c(
  "Account Control",
  "Action Financial Services",
  "Central Research",
  "Coast Professional",
  "Collecto",
  "Credit Adjustments",
  "FMS",
  "GC Services",
  "Immediate Credit Recovery",
  "National Recoveries",
  "Pioneer",
  "Windham")) {
    loans$agency_name[grepl(agency_i,loans$agency_name)]<-agency_i 
}

# Replace NA values with zeroes for calculations
loans$rehabilitation[is.na(loans$rehabilitation)] <- 0

# Make a new dataframe to compute each repayment type as a portion of total
loans_relative <- loans %>% 
  select(c(agency_name, total, consolidation:wage_garnishments)) %>% 
  transform(consolidation = consolidation/total,
            rehabilitation = rehabilitation/total,
            voluntary_payments = voluntary_payments/total,
            wage_garnishments = wage_garnishments/total)
  
# Create a summary dataframe organized by agency_name
loan_summary <- loans_relative %>% 
  group_by(agency_name) %>% 
  summarize(total = mean(total)/(10^6),
            consolidation = mean(consolidation),
            rehabilitation = mean(rehabilitation),
            voluntary_payments = mean(voluntary_payments),
            wage_garnishments = mean(wage_garnishments))

names(loan_summary)[3] <- paste("Consolidation")
names(loan_summary)[4] <- paste("Rehabilitation")
names(loan_summary)[5] <- paste("Voluntary Payments")
names(loan_summary)[6] <- paste("Wage Garnishments")
  
# Rearrange data for stacked column plot
bar_df <- loan_summary %>% 
  gather("Consolidation", "Rehabilitation", "Voluntary Payments", "Wage Garnishments",
         key = "source", value = "proportion") %>% 
  arrange(total)

# Generate a name vector for the column plot to include total repayment amount for agency
names_list <- NULL
for(nam in unique(bar_df$agency_name)){
  names_list <- 
    append(names_list, paste0(nam," ($",round(unique(bar_df$total[bar_df$agency_name == nam]),0),"M)"))
}

# Geneate column plot
column_plot <- ggplot(bar_df) +
  geom_col(mapping = aes(x = agency_name, y = proportion, fill = source)) +
  coord_flip() + xlab(NULL) + ylab("Portion of Total Repayments for Agency") +
  ggtitle("Portion of Repayments by Source", 
          subtitle = "Mean Total Repayments in Descending Order") +
  scale_fill_brewer(palette = "Accent") +
  scale_fill_discrete(name = "Payment Category") +
  scale_x_discrete(labels = names_list)
column_plot
```

While this is interesting if you are curious about each individual `Agency`, it does not give much insight into any over all trends. I was hoping that you may see a trend in repayment composition as a function of `Total Repayments`, which is why I sorted it that way; this doesn't seem to be the case.

## Repayment Category Proportion by Total Repayment

Next, let's look at how the proportion of each `Repayment Type` varies with `Total Repayment`. Each point represents an `Agency`.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align="center"}

# Generate faceted plot for all repayment types
facet_plot <- ggplot(bar_df, aes(total, proportion), labeller = labeller(supp = labs)) + 
  geom_point(alpha = 0.6) +
  facet_wrap(nrow = 2, ncol = 2, facets = vars(source), scales = "free") +
  theme(strip.background = element_rect(
    color="gray", fill="#90F0FA", size=1.5)) +
  ylab("Portion of Total Repayments") +
  xlab("Agency Mean Total Repayments (Millions of U.S. Dollars)") +
  ggtitle("Portion of Total Repayments by Category for Each Agency")
facet_plot
```

From this plot, we can see that there are some general trends between the `Repayment Type` proportions and `Total Repayment` amounts:

* For `Consolidation`, `Voluntary Payments`, and `Wage Garnishment`, proportion decreases with increasing `Total Repayment`
* Proportion `Rehabilitation` increases with increasing `Total Repayment`

The more interesting thing to note though, is that in all of the graphs there is a group of points that defy this trend, where proportion of `Consolidation`, `Voluntary`, and `Wage Garnishment` is low for low `Total Repayment`, and proportion `Rehabilitation` is high for low `Total Repayment`.

Upon examination however, the outlier groups do not consist of the same `Agency`s across plots. In fact, no single `Agency` is present in all four outlier groups. However, the outliers in all plots except `Wage Garnishment` consist of these 12 `Agency`s:

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align="center"}
outliers <- loan_summary %>% 
  filter(total < 100 & Consolidation < 0.15 & Rehabilitation > 0.65 & `Voluntary Payments` < 0.05)
  
outliers$agency_name %>% 
  matrix(3, byrow = TRUE) %>% 
  kable() %>% kable_styling()
```

## Covariation Between Repayment Categories

A final analysis that was investigated was how do the proportion of the different `Payment Types` vary together.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.height=6}

pairwise_plot <- loan_summary %>% select(-agency_name, -total) %>%
  ggpairs(
    upper = list(continuous = "cor", combo = "box_no_facet", discrete = "facetbar", na = "na"),
    diag = list(continuous = wrap("barDiag", bins = 12), discrete = "barDiag", na = "naDiag"))

pairwise_plot
```

From this analysis two main observations were made:

* `Wage Garnishment` does not strongly correlate with any of the other payment types
* `Consolidation`, `Rehabilitation`, and `Voluntary Payments` do covary, with the following relationships:
  * `Consolidation` and `Voluntary Payments` are negatively correlated with `Rehabilitation` (r<sup>2</sup> = -0.948 & r<sup>2</sup> = -0.946)
  * `Consolidation` is positively correlated with `Voluntary Payments` (r<sup>2</sup> = 0.953)
  
# Conclusions

* Overall repayment category composition for each `Agency` does not depend on `Total Repayment` amount.
* `Rehabilitation` is positively correlated with `Total Repayment`, while all other categories are negatively correlated.
* All repayment types co-vary with strong correlations, with the exception of `Wage Garnishments`.

# Authorship

Kenneth Schackart, [\@SchackartK](https://twitter.com/SchackartK)